Bayesian exponential random graph modeling of KWL Burial network

### 2020-07-17 analysis/paper/paper.Rmd
update code and edit the method section

```{r tidy-data}
# combine some periods
burial_three_period_tidy <-
  burial %>%
  rename(burial_label = ID) %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>%
  mutate(Gold_leaf = ifelse(Gold_leaf == "shatter", "1", Gold_leaf),
         Stoneware = ifelse(Stoneware == "base", "1", Stoneware),
         Stamped_ceramic = ifelse(Stamped_ceramic == "cluster", "1", Stamped_ceramic)) %>%
  mutate_at(21:ncol(.), as.numeric) %>%
  janitor::remove_empty(which = "cols") %>%
  mutate(total = rowSums(.[c(21:48, 50, 55, 56)], na.rm = TRUE)) %>% #prestige goods
  mutate(Porcelain = rowSums(.[c(40:48, 55, 56)], na.rm = TRUE)) %>% #B&W, porcelain, Anping, kendi
  mutate(Porcelain = ifelse(Porcelain == 0, NA, Porcelain)) %>%
  mutate(quantity = case_when(
    total == 0 ~ "none",
    total > 0 & total <= 7 ~ "low",
    total > 7 & total < 100 ~ "medium",
    total >= 100 ~ "high",
    TRUE ~ "other")) %>% # the classification is based on the result of histogram
  mutate(Age_scale = case_when(
    `Age` %in% c("1","2") ~ "0-12",
    `Age` == "3" ~ "12~20",
    `Age` %in% c("4","5","6","7","8") ~ "+20",
    TRUE ~ "NA")) %>%
    mutate(gender = case_when(
    `Gender` %in% c("1","2") ~ "male",
    `Gender` %in% c("3","4") ~ "female",
    TRUE ~ "NA")) %>%
  mutate(Gold_bead_low = ifelse(Golden_bead == 1, 1, NA),
         Gold_bead_med = ifelse(Golden_bead > 1 & Golden_bead <10, 1, NA),
         Gold_bead_high = ifelse(Golden_bead > 10, 1, NA),
         Agate_bead_low = ifelse(Agate_bead == 1, 1, NA),
         Agate_bead_med = ifelse(Agate_bead > 1 & Agate_bead <10, 1, NA),
         Agate_bead_high = ifelse(Agate_bead > 10, 1, NA),
         `Indo-Pacific_bead_low` = ifelse(`Indo-Pacific_bead` < 100, 1, NA),
         `Indo-Pacific_bead_med` = ifelse(`Indo-Pacific_bead` > 100 & `Indo-Pacific_bead` < 900, 1, NA),
         `Indo-Pacific_bead_high` = ifelse(`Indo-Pacific_bead` > 900, 1, NA)) %>% 
  # the low-med-high index is determined by interpretation for histograms for those burial good types
  select(burial_label,
         Phase,
         Age_scale,
         gender,
         Gold_bead_low,
         Gold_bead_med,
         Gold_bead_high,
         Agate_bead_low,
         Agate_bead_med,
         Agate_bead_high,
         #Agate_bead, #female burials
         #Golden_bead,
         Porcelain,
         Gold_leaf, #prestige good
         fish_shape_knit, #prestige good
         #Bell, #children's burials
         quantity)
         #total) # select specific variable to drop columns (uninformative variables)

# number of each phase
burial_three_period_number <-
  burial_three_period_tidy %>%
  count(Phase)
```

@@

<!--chronology-->
We assigned burials to the pre-European period (n = 26), European and post-European period (n = 45) for network comparison according to depth, radiocarbon dates, diagnostic materials, stratigraphic data, and previous studies. Since the sedimentary characteristics show vertical and lateral continuity cross units of AD section, depth is used as the first criterion for burial chronology. The chronology is then evaluated by the burials with time marker, light gray glazed jars, also called An-ping jars. The jars were believed to be produced in Southeast China and circulated in Southeast Asia from the late 16th to 18th century, but are often found at sites with the European occupation in the early 17th century in Taiwan [@Berrocal2018; @Hsieh1995; @Ketel2011]. In addition to the diagnostic jars, three radiocarbon dates of charcoal associated with three different burials are used as another evidence to evaluate the chronology [@Chen2007]. Some burials that were heavily disturbed by modern construction are excluded to avoid misinterpreting due to incomplete data. With the information of burials, including quantity of grave goods, types of grave goods, burial identities, we can reconstruct social relations between individuals by linking them according to similar prestige goods. This is based on the assumption that the prestige goods imply social practices of trade, exchange, and gifting that reflect social relations between individuals [@Coward2013, pp. 252]. Also, prestige materials usually relate to a representation of social status, where the same economic or social class, tend to use similar goods as a status indicators [citations]. Therefore, we build network where burials represent actors (nodes in the network) that are linked when they have th same prestige goods in common. The prestige goods we identified including agate beads, gold-foil beads, imported Chinese porcelains, gold foils, and fish-shaped ornaments, according to previous research [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011] and the description in historical records [@Borao2009; @LiandWu2006].

@@

We hypothesized that after the start of the European presence in the 17th century, a group of ambitious individuals competed with each other for access to trade with the Europeans due to the high value of trade goods in their society. The high value of trade goods had been identified by the historical documents written by the Europeans and the Han Chinese [@Borao2009; @LiandWu2006]. In addition to valuable foreign goods, the indirect colonialism in northeastern Taiwan may suggest the potentially manipulation of the European colonial image by local ambitious individuals for building their personal status and power. This led to the monopolization of valuable goods among a small numbers of individuals and the society gradually switched to a structure with more social inequality. If the social inequality gradually increased in Kiwulan, then we will observe it from the burials by analyzing the networks generated by those social processes. Network analysis provides a way to understand the underlying mechanisms that shape the relationships between people at Kiwulan, and the effect of foreign contact on these relationships. Our purpose is to identify changes in network structure of burials before and after the foreign contacts, and explore which variables may have contributed to the observed economic inequality across burials using network analysis approaches. This is important to understand how an Indigenous society reacted to the presence of foreign imperial power.

@@

MCMC simulation can estimate the posterior distribution of interest by of random sampling without the need of normal distribution for the prior and likelihood. MCMC is a random probability distribution that describes a sequence of events or transitions (chain) from one state to another state following probabilistic rules. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and than switch to a pair of actors at random within that chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000. The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. This provides a statistical way to check how well the estimated posterior parameter distribution can reproduce networks similar to the general structural features of the observed network. Our observed networks, pre-European and post-European networks, followed the normal distribution that is a suitable prior model in network analysis that assumes networks with low density and high transitivity commonly observed in real world [@Caimo2017]. Since our original data implies higher transitivity due to potentially multiple groups having the same kind of burial good in common, we specified the prior based on our observation with higher transitivity.

>>>>

```{r tidy-data}
# combine some periods
burial_three_period_age_tidy <-
  burial %>%
  rename(burial_label = ID) %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>%
  mutate(Gold_leaf = ifelse(Gold_leaf == "shatter", "1", Gold_leaf),
         Stoneware = ifelse(Stoneware == "base", "1", Stoneware),
         Stamped_ceramic = ifelse(Stamped_ceramic == "cluster", "1", Stamped_ceramic)) %>%
  mutate_at(21:ncol(.), as.numeric) %>%
  janitor::remove_empty(which = "cols") %>%
  mutate(total = rowSums(.[c(21:48, 50, 55, 56)], na.rm = TRUE)) %>% #prestige goods
  mutate(Porcelain = rowSums(.[c(40:48, 55, 56)], na.rm = TRUE)) %>% #B&W, porcelain, Anping, kendi
  mutate(Porcelain = ifelse(Porcelain == 0, NA, Porcelain)) %>%
  mutate(quantity = case_when(
    total == 0 ~ "none",
    total > 0 & total <= 7 ~ "low",
    total > 7 & total < 100 ~ "medium",
    total >= 100 ~ "high",
    TRUE ~ "")) %>% # the classification is based on the result of histogram
  mutate(Age_scale = case_when(
    `Age` %in% c("1","2") ~ "0-12",
    `Age` == "3" ~ "12~20",
    `Age` %in% c("4","5","6","7","8") ~ "+20",
    TRUE ~ "NA")) %>%
  mutate(gender = case_when(
    `Gender` %in% c("1","2") ~ "male",
    `Gender` %in% c("3","4") ~ "female",
    TRUE ~ "NA")) %>%
  mutate(ritual = case_when(
    `Stamped_ceramic` == "2" ~ "two pots",
    `Stamped_ceramic` == "1" ~ "one pot",
    TRUE ~ "NA")) %>%
  mutate(Gold_bead_low = ifelse(Golden_bead == 1, 1, NA),
         Gold_bead_med = ifelse(Golden_bead > 1 & Golden_bead <10, 1, NA),
         Gold_bead_high = ifelse(Golden_bead > 10, 1, NA),
         Agate_bead_low = ifelse(Agate_bead == 1, 1, NA),
         Agate_bead_med = ifelse(Agate_bead > 1 & Agate_bead <10, 1, NA),
         Agate_bead_high = ifelse(Agate_bead > 10, 1, NA),
         `Indo-Pacific_bead_low` = ifelse(`Indo-Pacific_bead` < 100, 1, NA),
         `Indo-Pacific_bead_med` = ifelse(`Indo-Pacific_bead` > 100 & `Indo-Pacific_bead` < 900, 1, NA),
         `Indo-Pacific_bead_high` = ifelse(`Indo-Pacific_bead` > 900, 1, NA)) %>% #based on the result of histogram
  select(burial_label,
         Phase,
         Age_scale,
         gender,
         ritual, # consider to remove if not very informative
         Gold_bead_low,
         Gold_bead_med,
         Gold_bead_high,
         Agate_bead_low,
         Agate_bead_med,
         Agate_bead_high,
         Porcelain, #prestige good
         Gold_leaf, #prestige good
         fish_shape_knit, #prestige good
         quantity)

# number of each phase
burial_three_period_number <-
  burial_three_period_tidy %>%
  count(Phase)

pre_E <- burial_three_period_number$Phase[4]
post_E <- burial_three_period_number$Phase[3]

```

@@

<!--chronology-->
We assigned burials to the pre-European period (n = `r pre_E`), European and post-European period (n = `r post_E`) for network comparison according to depth, radiocarbon dates, diagnostic materials, stratigraphic data, and previous studies. Since the sedimentary characteristics show vertical and lateral continuity cross units of AD section, depth is used as the first criterion for burial chronology. The chronology is then evaluated by the burials with time marker, light gray glazed jars, also called An-ping jars. The jars were believed to be produced in Southeast China and circulated in Southeast Asia from the late 16th to 18th century, but are often found at sites with the European occupation in the early 17th century in Taiwan [@Berrocal2018; @Hsieh1995; @Ketel2011]. In addition to the diagnostic jars, three radiocarbon dates of charcoal associated with three different burials are used as another evidence to evaluate the chronology [@Chen2007]. Some burials that were heavily disturbed by modern construction are excluded to avoid misinterpreting due to incomplete data. With the information of burials, including quantity of grave goods, types of grave goods, burial identities, we can reconstruct social relations between individuals by linking them according to similar prestige goods. This is based on the assumption that the prestige goods imply social practices of trade, exchange, and gifting that reflect social relations between individuals [@Coward2013, pp. 252]. Also, prestige materials usually relate to a representation of social status, where the same economic or social class, tend to use similar goods as a status indicators [citations]. Therefore, we build network where burials represent actors (nodes in the network) that are linked when they have th same prestige goods in common. The prestige goods we identified including agate beads, gold-foil beads, imported Chinese porcelains, gold foils, and fish-shaped ornaments, according to previous research [@Cheng2008; @Hsieh2009; @Hsieh2012; @Wang2011] and the description in historical records [@Borao2009; @LiandWu2006].

@@

Our hypothesis is that after the start of the European presence in the 17th century, a group of ambitious individuals competed with each other for access to trade with the Europeans due to the high value of trade goods in their society. The high value of trade goods had been identified by the historical documents written by the Europeans and the Han Chinese [@Borao2009; @LiandWu2006]. In addition to valuable foreign goods, the indirect colonialism in northeastern Taiwan may suggest the potentially manipulation of the European colonial image by local ambitious individuals for building their personal status and power. This led to the monopolization of valuable goods among a small numbers of individuals and the society gradually switched to a structure with more social inequality. If the social inequality gradually increased in Kiwulan, then we will observe it from the burials by analyzing the networks generated by those social processes. Network analysis provides a way to understand the underlying mechanisms that shape the relationships between people at Kiwulan, and the effect of foreign contact on these relationships. Our purpose is to identify changes in network structure of burials before and after the foreign contacts, and explore which variables may have contributed to the observed economic inequality across burials using network analysis approaches. This is important to understand how an Indigenous society reacted to the presence of foreign imperial power.

@@

Markov chain Monte Carlo (MCMC) is an estimation of posterior distribution of interest through directly random sampling the posterior without assuming the prior is a normal distribution. We can obtain a distribution by constructing a Markov chain that describes a sequence of moves from current state to the next state following probabilistic rules based on algorithms. This enables a random or stochastic simulation in a long run without depending on the previous move. More chains ensures a more desirable distribution close to the target distribution that we seek, which is called the convergence of the Markov chains. In Bayesian ERGMs, MCMC first selects a set of edges (or a set of empty pair of actors) with equal probability, and than switch to a pair of actors at random within that chosen set [@Caimo2011]. In our case, we set the number of chains to six. For each chain, the number of burn-in iterations is 200 and the number of iterations after the burn-in is 2000. We set the number of iterations used to simulate a network y′ at each iteration to 10,000. The models are assessed by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distribution of the parameters of the model [@Caimo2011; @Caimo2017]. This provides a statistical approach to check how well the estimated posterior parameter distribution can reproduce networks similar to the general structural features of the observed network. Our observed networks, pre-European and post-European networks, followed the normal distribution that is a suitable prior model in network analysis that assumes networks with low density and high transitivity commonly observed in real world [@Caimo2017]. Since our original data implies higher transitivity due to potentially multiple groups having the same kind of burial good in common, we specified the prior based on our observation with high transitivity and low density.
